---
title: "HW3_613_Sun"
author: "Mike Sun"
date: "3/22/2022"
output: pdf_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
dw = "~/Desktop/HW3/Data/"

datjss <- read.csv(paste(dw,"datjss.csv",sep=""))
datsss <- read.csv(paste(dw,"datsss.csv",sep=""))
datstu_v2 <- read.csv(paste(dw,"datstu_v2.csv",sep=""))

datsss_nna = na.omit(datsss)
```

```{r,include=FALSE}
library(dplyr)
library(tidyverse)
```


$$Exercise 1$$
Q1: Number of students, schools, programs

Students:
```{r}
student_num = length(datstu_v2$V1)
student_num
```

Schools:
```{r}
school_num = length(unique(datsss_nna$schoolcode))
school_num
```

Programs:

```{r}
program <- subset(datstu_v2,select =
                        c(choicepgm1,choicepgm2,choicepgm3,choicepgm4,choicepgm5,choicepgm6))
program <- program %>% 
  pivot_longer(cols = starts_with("choicepgm"),names_to = "program_rank",values_to = "choiceprogram")

program_num = length(unique(program$choiceprogram))

program_num
```

Q2: Number of choices (school, program)

```{r}
school <- subset(datstu_v2,select =
                        c(V1,schoolcode1,schoolcode2,schoolcode3,schoolcode4,schoolcode5,schoolcode6))
program <- subset(datstu_v2,select =
                        c(choicepgm1,choicepgm2,choicepgm3,choicepgm4,choicepgm5,choicepgm6))

school <- school %>% 
  pivot_longer(cols = starts_with("schoolcode"),names_to = "school_rank",values_to = "schoolcode") 

program <- program %>% 
  pivot_longer(cols = starts_with("choicepgm"),names_to = "program_rank",values_to = "choiceprogram") 

school_program_matrix <- cbind(school,program) %>% select(schoolcode,choiceprogram) %>% unique()
```

```{r}
length(school_program_matrix$schoolcode)
```


Q3: Number of students applying to at least one senior high schools in the same district to home:

```{r}
school_district <- subset(datstu_v2,select =
                        c(V1,schoolcode1,schoolcode2,schoolcode3,schoolcode4,
                          schoolcode5,schoolcode6,jssdistrict))

school_district <- school_district %>% pivot_longer(
  cols = starts_with("schoolcode"),names_to = "schoolrank",values_to = "schoolcode")

district = na.omit(datsss_nna %>% select(schoolcode,sssdistrict) %>% unique())

school_district = left_join(school_district,district,by = "schoolcode")
```

```{r}
district_check = school_district %>%
  mutate(district_same = ifelse(jssdistrict == sssdistrict,1,0)) %>%
  select(V1,district_same) %>%
  unique()
```

```{r}
apply_sch_district = subset(district_check, district_same == 1)
length(apply_sch_district$V1)
```

Q4,5,6: Number of students each senior high school admitted, the cutoff of senior high schools, and the quality of senior high schools:

```{r}
school_rs <- subset(datstu_v2,select =
                        c(V1,schoolcode1,schoolcode2,schoolcode3,schoolcode4,
                          schoolcode5,schoolcode6,score,rankplace))

school_rs = subset(subset(school_rs,is.na(rankplace) == FALSE),is.na(score) == FALSE)
                     
school_rs <- school_rs %>% pivot_longer(
  cols = starts_with("schoolcode"),names_to = "schoolrank",values_to = "schoolcode")

school_rs <- school_rs %>% 
  mutate(schoolrank_num = ifelse(schoolrank == "schoolcode1",1,
                                 ifelse(schoolrank == "schoolcode2",2,
                                        ifelse(schoolrank == "schoolcode3",3,
                                               ifelse(schoolrank == "schoolcode4",4,
                                                      ifelse(schoolrank == "schoolcode5",5,6))))))

school_rs$rankplace = as.numeric(school_rs$rankplace)
school_rs$score = as.numeric(school_rs$score)
```


```{r}
school_rs_match = subset(school_rs,rankplace == schoolrank_num)

## admission
school_rs_final = school_rs_match %>% group_by(schoolcode) %>%
  mutate(school_admission_count = n()) %>% 
  mutate(ave_score = mean(score)) %>% 
  mutate(min_score = min(score)) %>% 
  select(schoolcode,school_admission_count,min_score,ave_score) %>%
  unique()

head(school_rs_final)
```

\newpage

$$Exercise 2$$
Create a school level dataset, where each row corresponds to a (school,program) with the following variables:

Q1&2: The district where the school is located, and the latitude and longitude of the district:

```{r}
datsss_nna_new = subset(datsss_nna,select=
                          c(schoolcode,sssdistrict, ssslong, ssslat)) %>% unique()

sch_prog = left_join(unique(school_program_matrix),datsss_nna_new,by="schoolcode")
```

Q3,4,5: cutoff (the lowest score to be admitted), quality (the average score of the students admitted), and size (number of students admitted)


```{r}
school_matrix_full <- datstu_v2 %>% 
  pivot_longer(cols = starts_with("schoolcode"),names_to = "school_rank",values_to = "schoolcode") 

program_matrix_full <- datstu_v2 %>% 
  pivot_longer(cols = starts_with("choicepgm"),names_to = "program_rank",values_to = "choiceprogram") %>%
  select(program_rank,choiceprogram)

sp_bind = cbind(school_matrix_full,program_matrix_full)
sp_full = subset(sp_bind,select =
  -c(choicepgm1,choicepgm2,choicepgm3,choicepgm4,choicepgm5,choicepgm6))
```

```{r}
sp_rs = subset(sp_full,select = 
                 c(score,rankplace,school_rank,program_rank,schoolcode,choiceprogram))

sp_rs = subset(subset(sp_rs,is.na(rankplace) == FALSE),is.na(score) == FALSE)

sp_rs <- sp_rs %>% 
  mutate(schoolrank_num = ifelse(school_rank == "schoolcode1",1,
                                 ifelse(school_rank == "schoolcode2",2,
                                        ifelse(school_rank == "schoolcode3",3,
                                               ifelse(school_rank == "schoolcode4",4,
                                                      ifelse(school_rank == "schoolcode5",5,6)))))) %>%
  mutate(programrank_num = ifelse(program_rank == "choicepgm1",1,
                                 ifelse(program_rank == "choicepgm2",2,
                                        ifelse(program_rank == "choicepgm3",3,
                                               ifelse(program_rank == "choicepgm4",4,
                                                      ifelse(program_rank == "choicepgm5",5,6))))))

sp_rs$rankplace = as.numeric(sp_rs$rankplace)
sp_rs$score = as.numeric(sp_rs$score)
```

```{r,warning=FALSE}
sp_rs_match = subset(sp_rs,rankplace == schoolrank_num)

## admission
sp_rs_final = sp_rs_match %>% group_by(schoolcode,choiceprogram) %>%
  mutate(school_admission_count = n()) %>% 
  mutate(ave_score = mean(score)) %>% 
  mutate(min_score = min(score)) %>% 
  select(choiceprogram,schoolcode,school_admission_count,min_score,ave_score) %>%
  unique()

head(sp_rs_final)
```

\newpage

$$Exercise 3$$
Distance:

```{r}
distance_school = left_join(sp_full, datsss_nna_new,by="schoolcode")
datjss_clean = subset(datjss,select = -c(X))
distance_school = left_join(distance_school,datjss_clean,by="jssdistrict")

distance_school = distance_school %>% 
  mutate(distance =
           sqrt((69.172*(ssslong-point_x)*cos(point_y/57.3))^2+(69.172*(ssslat-point_y))^2)
         )
```

```{r}
head(distance_school$distance)
```

\newpage

$$Exercise 4$$
Q1: Recode the schoolcode into its first three digits

```{r}
sp_full = sp_full %>% 
  mutate(scode_rev = substr(sp_full$schoolcode,1,3))

sp_full$scode_rev = as.numeric(sp_full$scode_rev)
```

Q2: Recode the program variable into 4 categories: arts (general arts and visual arts), economics (business and home economics), science (general science) and others

```{r}
sp_full = sp_full %>% 
  mutate(pgm_rev = 
           ifelse(choiceprogram == "General Arts","arts",
                ifelse(choiceprogram == "Visual Arts","arts",
                    ifelse(choiceprogram == "Business","economics",
                           ifelse(choiceprogram == "Home Economics","economics",
                                  ifelse(choiceprogram == "General Science","science","others")))))) 
```


Q3: Create a new choice variable choice rev

```{r}
sp_full = sp_full %>% 
  mutate(choice_rev = paste(scode_rev,pgm_rev,sep=""))
```

Q4: Recalculate the cutoff and the quality for each recoded choice

```{r}
sp_full <- sp_full %>% 
  mutate(schoolrank_num = ifelse(school_rank == "schoolcode1",1,
                                 ifelse(school_rank == "schoolcode2",2,
                                        ifelse(school_rank == "schoolcode3",3,
                                               ifelse(school_rank == "schoolcode4",4,
                                                      ifelse(school_rank == "schoolcode5",5,6)))))) %>%
  mutate(programrank_num = ifelse(program_rank == "choicepgm1",1,
                                 ifelse(program_rank == "choicepgm2",2,
                                        ifelse(program_rank == "choicepgm3",3,
                                               ifelse(program_rank == "choicepgm4",4,
                                                      ifelse(program_rank == "choicepgm5",5,6))))))

sp_full$rankplace = as.numeric(sp_full$rankplace)
sp_full$score = as.numeric(sp_full$score)
```

```{r}
sp_full_match = subset(sp_full,is.na(rank)==FALSE)
sp_full_match = subset(sp_full,rankplace == schoolrank_num)


## cutoff and quality
sp_full_final = sp_full_match %>% group_by(choice_rev) %>%
  mutate(ave_score = mean(score)) %>% 
  mutate(min_score = min(score)) %>% 
  select(V1,choice_rev,min_score,ave_score)

sp_full_unique = sp_full_final %>% select(choice_rev,min_score,ave_score) %>% unique()

head(sp_full_unique)
```

Q5: Consider the 20,000 highest score students

```{r}
high_score <- unique(subset(sp_full, select=c(V1,score)))
```

```{r}
topscore =arrange(high_score,desc(score))
topscore = topscore$V1[1:20000]

top_sp = sp_full %>% filter(V1%in%topscore)

top_sp_full = left_join(top_sp,sp_full_unique,by=c("choice_rev")) 

head(top_sp_full)
```

\newpage

$$Exercise 5$$

Propose a model specification. Write the Likelihood function:

I believe conditional logit serves better since we want to explore scores effect on each of school-program choice's probability instead of relative probability of other programs to one program (mlogit concept):

```{r}
top_sp_first <- top_sp_full %>% filter(schoolrank_num== 1) %>%
  mutate(choice = factor(choice_rev))
```

```{r}
like_fun = function(param)
{
  data = top_sp_first
  score =  data$score
  choice = data$choice

  ni = nrow(data)
  nj = length(unique(choice))
  ut = mat.or.vec(ni,nj)
  
  for (j in 1:nj)
  {
    # conditional logit
    ut[,j] = param[1] + param[2]*score[j]
  }
  prob   = exp(ut)
  prob   = sweep(prob,MARGIN=1,FUN="/",STATS=rowSums(prob))
  probc = prob[,1]
  
  probc[probc>0.999999] = 0.999999
  probc[probc<0.000001] = 0.000001
  like = sum(log(probc))
  return(-like)
}
```

testing:

```{r}
npar = 2
param = runif(npar)
like_fun(param)
```

Optimization:

```{r}
optim(par = c(-0.5,-0.2), fn=like_fun,method="BFGS")$par
```

By optimization results, we have beta equals -0.02. I fail to compute the marginal effect of beta obtained here.


$$Exercise 6$$



Propose a model specification. Write the Likelihood function:

Again, conditional logit will help us explore the relationship better, since we want to explore school quality (average admission effect) on each of school-program choice's probability instead of relative probability of other programs to one program:

```{r}
like_fun_2 = function(param)
{
  data = top_sp_first
  ave_score =  data$ave_score
  choice = data$choice

  ni = nrow(data)
  nj = length(unique(choice))
  ut = mat.or.vec(ni,nj)
  
  for (j in 1:nj)
  {
    # conditional logit
    ut[,j] = param[1] + param[2]*ave_score[j]
  }
  prob   = exp(ut)
  prob   = sweep(prob,MARGIN=1,FUN="/",STATS=rowSums(prob))
  probc = prob[,1]
  
  probc[probc>0.999999] = 0.999999
  probc[probc<0.000001] = 0.000001
  like = sum(log(probc))
  return(-like)
}
```


Optimization:

```{r}
optim(par = c(0.5,-0.1), fn=like_fun_2)$par
```

By optimization results, we have beta equals -0.14. I fail to compute the marginal effect of beta obtained here.

\newpage

$$Exercise 7$$

In this exercise, we are interested in the effect of excluding choices where the program is “Others”.

Q1.Explain and justify, which model (first or second model) you think is appropriate to conduct this ex- ercise:

Answer:

Since we are excluding "Others" program, school quality become less of a program and student test score now has a more direct impact of the three program. Previously, good school and bad school can have different preference over opening programs thus school quality would better reflect that changes. Now, since we are dealing with three major programs, school's quality become less of a concern, and student test score will show more of their preferences over different school-programs. Thus, first model is more appropriate.



